Synthesizing Quotient Lenses

### 2018-03-10 papers/QuotientLenses/main.tex
In progress edits to background/example section

\section{QRE Lenses by Example}
\label{sec:example}
\subsection{Bijective String Lenses}
Consider a bidirectional transformation that converts \bibtex{} citation records
such as
\begin{verbatim}
@Book {Lovelace,
Author = "Ada Lovelace",
Title = {Generic Title}
}
\end{verbatim}
\noindent
into an equivalent EndNote format like the following
\begin{verbatim}
%0 Book
%F Lovelace
%A Ada Lovelace
%T Generic Title
\end{verbatim}
\noindent
and vice versa.

Boomerang's bijective string lenses are designed to define a bidirectional
transformation such as this, where the data formats can be matched in a
one-to-one manner, in this case by matching the label, author and title fields.
Boomerang allows programmers to define such bijecive transformations using
combinators which favour a compositional approach in which the programmer
defines small lenses which are then put together using the combinators to
define larger and larger lenses until the final lens is defined.

Each bijective string lens is constructed from simple primitive combinators including:
\begin{itemize}
  \item \kw{del} $s$: \textit{delete} the constant string $s$ in the get
  direction; insert it in the put direction.
  \item \kw{ins} $s$: \textit{insert} the constant string $s$ in the get
  direction; delete it in the put direction.
  \item \kw{copy} $R$: copy the text matching the regular expression $R$ in
  both directions.
\end{itemize}
In addition, such primitives may be combined using the regular operators
Kleene star, alternation and concatenation, or functional composition of
lenses.  Figure~\ref{fig:example-lens} illustrates the use of these
combinators to transform \bibtex{} to EndNote (and vice versa).
\begin{figure}[t]
\begin{lstlisting}
let preamble : (lens in "@book{" $\Leftrightarrow$ "%0 Book\n%F ") =
del "@book{" . ins "%0 Book\n%F "

@@

let bib_to_end : (lens in bibtex $\Leftrightarrow$ end_note) = preamble . label . author_lens . title_lens
\end{lstlisting}
\caption{An example lens between \cd{bibtex} and \cd{end_note} regular
expressions with no equivalences.\bcp{And a readability suggestion:
define a named constant for a quote character.}}
\label{fig:example-lens}
\end{figure}

Unfortunately, programming bijective string lenses in Boomerang is often slow and
tediuos because of the fiddly constraints that Boomerang imposes on its
terms. making lens programming slow and tedious. For example, Boomerang
programmers often must rearrange the order of data items by recursively using
operators that \swap adjacent fields. Furthermore, the Boomerang type checker is
very strict, disallowing many programs because they contain ambiguity about how
certain data is transformed.

In our previous work in~\cite{optician}, we presented a solution to these
shortcomings of bijective string lenses by developing a tool called {\em Optician}
which enables a programmer to synthesize a bijective string lens from a pair of regular
expressions $(S, T)$ and a set of input-output example \cd{exs} pairs that
determines how the synthesized lens should behave on those examples. With {\em
Optician}, the programmer need only spend the effort of specifying the data
formats and the behavior of the lens on a few inputs in order to define a
bijective string lens. As a result defining bijective string lenses becomes faster and less
error prone, since the programmer need only invoke the synthesis procedure
\lstinline{synth S <=> T using exs} in order to define a bijective string lens rather
than manually derive the lens herself.

\subsection{Quotient Lenses}
Unfortunately, while our synthesis tool greatly simplifies the task of
programming bijective string lenses, not all bidirectional tranformations are bijective
in nature. For instance, users are typically uninterested in preserving
whitespace between words or between fields.  The order of author and title
fields is also likely irrelevant and there may be equivalent ways of writing
the same name:  ``Lovelace, Adan'' vs ``Ada Lovelace.''

When two instances of the same format differ only terms of such inessential
information, it is useful\bcp{Would it be worth spelling out the alternative
(i.e., what lenses would look like if they had to preserve all this stuff?)}
to treat them as identical and to map them into canonical representations.
Continuing our example, the following two \bibtex{} citations represent the same
logical object and as a result, should be translated to the same EndNote record.
\begin{verbatim}
@Book {Lovelace,
Author = "Ada Lovelace",
Title = {Generic Title},
}

@@

\begin{theorem}\label{normal form}
If there is a derivation $q : Q_1 \Leftrightarrow Q_2$, then there exists a
bijective lens $\ell : K(Q_1) \Leftrightarrow K(Q_2)$ such that:
\begin{align*}
\llbracket q \rrbracket.\get &= \llbracket \ell \rrbracket\circ
\canonizer(Q_1)\\
\llbracket q \rrbracket.\lput &= \llbracket \ell \rrbracket^{-1} \circ
\canonizer(Q_2)
\end{align*}
\end{theorem}
\begin{proof}
By straightforward induction on the derivation
$q_1 : Q_1 \Leftrightarrow Q_2$. Since bijective lenses are closed under strongly unambiguous
regular operators, the tricky case is composition.  The composition case is
more involved since we need to show that we can get rid of the canonizers in
the middle:

\sam{TODO: Put a diagram here to show what you mean.}

>>>>

\section{Background: Bijective String Languages}
\label{sec:background}
Before describing QREs and QRE lenses, we briefly review bijective
string lenses. Consider a bidirectional transformation that converts
between \bibtex{} citation records such as
\begin{verbatim}
  @Book {Lovelace,
  Author = "Ada Lovelace",
  Title = {Notes}
}
\end{verbatim}
\noindent
and equivalent EndNote records like the following
\begin{verbatim}
  %0 Book
  %F Lovelace
  %A Ada Lovelace
  %T Notes
\end{verbatim}
\noindent
Boomerang's bijective string lenses are designed to define bidirectional
transformations such as this one, where the data formats can be matched in a
one-to-one manner, in this case by matching the label, author and title fields.
Boomerang encourages a compositional approach in which programmers
define simple lenses and then compose them using a variety of
combinators.

Primitive lenses include:
\begin{itemize}
  \item \kw{del} $s$: delete the constant string $s$ in the get
  direction; insert it in the put direction.
  \item \kw{ins} $s$: insert the constant string $s$ in the get
  direction; delete it in the put direction.
  \item \kw{copy} $R$: copy the text matching the regular expression $R$ in
  both directions.
\end{itemize}
Such primitives may be combined using the regular operators
Kleene star, alternation and concatenation, as well as lens composition.
Figure~\ref{fig:example-lens} illustrates the use of these
combinators to define a lens \cd{bib_to_end} that transforms data
in \bibtex{} to EndNote (and vice versa). 
\begin{figure}[t]
\begin{lstlisting}
let preamble : (lens in "@book{" $\Leftrightarrow$ "%0 Book\n%F ") =
del "@book{" . ins "%0 Book\n%F "

@@

let bib_to_end : (lens in bibtex $\Leftrightarrow$ end_note) = preamble . label . author_lens . title_lens
\end{lstlisting}
\caption{An example bijective lens \cd{bib_to_end} between data
matching \cd{bibtex} and \cd{end_note} regular 
expressions, which are omitted for brevity.  
\bcp{And a readability suggestion:
define a named constant for a quote character.}}
\label{fig:example-lens}
\end{figure}

Recent work described the {\em Optician}
algorithm/tool~\cite{optician}, which can synthesize bijective lenses
such as \cd{bib_to_end}, obviating the sometimes tedious tasks
involved in writing such lenses by hand.  Given the directive
\begin{lstlisting}
synth S <=> T using exs
\end{lstlisting}
\noindent
{\em Optician} will synthesize a bijective lens between 
source and target formats described by regular expressions
\cd{S} and \cd{T}, respectively, and a set of input-output example pairs \cd{exs} specifying how
the synthesized lens should behave on those examples.

\section{QRE Lenses by Example}
\label{sec:example}

\subsection{Quotient Lenses}
Unfortunately, while {\em Optician} greatly simplifies the task of
programming bijective string lenses, not all bidirectional tranformations are bijective
in nature. For instance, users are typically uninterested in preserving
whitespace between words or between fields.  The order of author and title
fields is also likely irrelevant and there may be equivalent ways of writing
the same name:  ``Lovelace, Ada'' vs ``Ada Lovelace.''

When two instances of the same format differ only terms of such inessential
information, it is desirable
to treat them as identical and to map them into canonical
representations.
The alternative is to require extra machinery to preserve the
inessential details, which leads to complex and brittle lenses.
Continuing our example, the following two \bibtex{} citations represent the same
logical object and as a result, should be translated to the same EndNote record.
\begin{verbatim}
@Book {Lovelace,
Author = "Ada Lovelace",
Title = {Generic Title},
}

@@

\begin{theorem}\label{normal form}
If there is a derivation $q : Q_1 \Leftrightarrow Q_2$, then there exists a
bijective lens $\ell : K(Q_1) \Leftrightarrow K(Q_2)$ such that:
\begin{align*}
\llbracket q \rrbracket.\get &= \llbracket \ell \rrbracket\circ
\canonizer(Q_1)\\
\llbracket q \rrbracket.\lput &= \llbracket \ell \rrbracket^{-1} \circ
\canonizer(Q_2)
\end{align*}
\end{theorem}
\begin{proof}
The proof follows by induction on the derivation of $q : c \Leftrightarrow c$.
The most interesting part of the proof is the case for functional composition
as we must demonstrate that it is possible to eliminate the canonizers in the
middle of the term.
